# Data

## Description

The project aims to analyze affordable housing units in NYC. The dataset originates from The Department of Housing Preservation and Development (HPD), a department of the government of NYC dedicated to ensuring quality and affordable housing while sustaining neighborhood strength and diversity.  

The dataset can be easily accessed by downloading the CSV file from https://data.cityofnewyork.us/Housing-Development/Affordable-Housing-Production-by-Building/hg8x-zxpr/about_data, which can be directed via the NYC Housing Preservation & Development government website (https://www.nyc.gov/site/hpd/about/open-data.page). 

The dataset was compiled by HPD using various methods, including wage information for housing development projects as mandated by Local Law 44, registration details from owners of residential rental properties, and survey data from housing units across the city. It consists of 41 columns and 7,637 rows, with each row representing details of a housing development or preservation project. According to the HPD, the dataset was first published to the public on December 20, 2016. It undergoes quarterly updates and the most recent update occurred on September 23, 2024.

One notable issue with the dataset is that some entries in the Project Completion Date column remain empty, indicating that the project is still ongoing. Therefore, it is necessary to manually input NA values to these empty entries.

```{r, include = FALSE, message = FALSE, warning = FALSE}
library(readr)
library(dplyr)
data <- read.csv("/Users/chrisyang/Downloads/Housing.csv")
```

```{r}
colnames(data)
```

Data Explanation 

This project does not require every column from the dataset, as its focus is on analyzing affordable housing units in NYC. Outlined below are the selected columns from the dataset that may be utilized for the project.

Data: Affordable Housing Production by Building

1. Project.ID: unique numeric identifier assigned to each project by HPD.

2. Project Start Date: date of the project loan or agreement closing.

3. Project Completion Date: date that the last building in the project was completed. If the project has not yet completed, then the field is blank.

4. Borough: borough where the building is located.

5. Community Board: the New York City Community District where the building is located.

6. Reporting.Construction.Type: indicates whether the building is categorized as ‘new construction’ or ‘preservation’ in Housing New York statistics.

7. Extremely Low Income Units: units with rents that are affordable to households earning 0 to 30% of the area median income 

8. Very Low Income Units: units with rents that are affordable to households earning 31 to 50% of the area median income

9. Low Income Units: units with rents that are affordable to households earning 51 to 80% of the area median income

10. Moderate Income Units: units with rents that are affordable to households earning 81 to 120% of the area median income

11. Middle Income Units: units with rents that are affordable to households earning 121 to 165% of the area median income

12. Other Income Units: units reserved for building superintendents

13. Studio Units: units with 0-bedrooms.

14. 1-BR Units: units with 1-bedroom.

15. 2-BR Units: units with 2-bedrooms.

16. 3-BR Units: units with 3-bedrooms.

17. 4-BR Units: units with 4-bedrooms.

18. 5-BR Units: units with 5-bedrooms.

19. 6-BR+ Units: units with 6-bedrooms or more.

20. Counted Rental Units: units in the building, counted toward the Housing New York plan, where assistance has been provided to landlords in exchange for a requirement for affordable units.

21. Counted Homeownership Units: units in the building, counted toward the Housing New York Plan, where assistance has been provided directly to homeowners.

22. Total Units: total number of units, affordable and market rate, in each building.

```{r, include = FALSE, message = FALSE, warning = FALSE}
data <- data |> 
  select("Project.ID", "Project.Start.Date", "Project.Completion.Date", "Borough", "Community.Board",
         "Extremely.Low.Income.Units", "Reporting.Construction.Type", "Very.Low.Income.Units", "Low.Income.Units",
         "Moderate.Income.Units", "Middle.Income.Units","Other.Income.Units", "Studio.Units","X1.BR.Units",
         "X2.BR.Units", "X3.BR.Units", "X4.BR.Units", "X5.BR.Units", "X6.BR..Units", "Counted.Rental.Units",
         "Counted.Homeownership.Units", "Total.Units")

data$Project.Completion.Date[data$Project.Completion.Date == ""] <- NA
```


## Missing value analysis

```{r, include = FALSE, message = FALSE, warning = FALSE}
library(tidyverse)
library(mi)
library(Matrix)
library(redav)
library(ggplot2)
```


To examine patterns in missing values, we begin by determining the count of NAs in each column.

```{r}
NA_per_column <- colSums(is.na(data)) |> sort(decreasing = TRUE)
NA_per_column
```
The Project.Completion.Date column has 1,633 missing entries.

```{r, warning = FALSE}
x <- missing_data.frame(data)
class(x)
```

```{r}
levels(x@patterns)
```

```{r}
summary(x@patterns)
```

```{r}
image(x)
```

To uncover potential patterns among these NAs, we analyzed the data and plotted a heatmap of them by utilizing the "mi" package, which is particularly useful for datasets with a large number of rows.

The dark areas of the figure represent missing values. The figure does not suggest a possible pattern related to data collection issues since there is only one column with NA values.

```{r}
plot_missing(data, max_cols = 3, percent = FALSE) 
```

Another visualization reveals the same pattern of the NA values. Approximately 6,000 rows have complete data across the columns, while around 1,600 rows have missing values in the Project Completion Date column.


```{r}
all_boroughs <- unique(data$Project.Completion.Date)

NA_grouping <- data |>
  group_by(Borough) |>
  summarize(
    num_rows = n(),
    NA_num = sum(is.na(Project.Completion.Date)),
    NA_percent = (NA_num / num_rows) * 100
  )

ggplot(NA_grouping, aes(x = Borough, y = NA_percent, fill = Borough)) +
  geom_bar(stat = "identity") +
  labs(
    title = "Percentage of NA Values in Project Completion Date Column by Borough",
    x = "Borough",
    y = "Percentage of NA Values"
  )
```

To determine whether missing values in the Project.Completion.Date column are correlated with values of another variable, we analyzed the percentage of NA values by borough (Bronx, Brooklyn, Manhattan, Queens, and Staten Island). The results reveal that Queens marked the highest rate of NA values (approximately 32%) while Staten Island had the lowest proportion (approximately 2-3%). The rest of the boroughs marked approximately equal rate of NA values at around 20%. This analysis may highlight a clear geographic trend in the missing data, with Queens emerging as the most problematic borough with ongoing projects and Staten Island being the least affected with most of projects completed. It might also suggest that the number of projects varies significantly between boroughs. 